Deep Gaussian Processes and Hierarchical Modeling with Functional Data: A Case Study in Cosmological Power Spectra

### 2024-09-20 paper/main.tex
update manuscript

Leveraging these simulations allows for emulation and prediction of the matter power spectrum under varying specifications for eight different cosmological parameters (i.e., different cosmologies). Estimating and understanding the influence of these parameters on cosmic expansion are areas of active research. In addition to presenting the Mira-Titan simulation suite, \cite{moran2023mira} builds off of previous spectrum emulation \citep{lawrence2017mira} to provide the final emulator (Cosmic Emu) based on the full suite of simulations.

In this work, we expand on the previous emulation methods and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical and basis functions modeling within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}. \textcolor{red}{Please note: this repository is currently private until we hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 117 different cosmologies. Each of these cosmologies is specified by 8 cosmological parameter values, which we will denote as $\theta \in \mathbb{R}^8$. Different simulation output exists for different values of a ninth value, the redshift; in this work, we fix this value at 0. For each cosmology, we have a set of 18 simulated spectra: an inexpensive spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from a high resolution simulation ($y_h$).

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_data.png}
    \caption{Left: The perturbation theory, low resolution runs, and high resolution run for the first training cosmology. The weighted average is also shown as a dashed line. Dotted lines at the bottom indicate where each data type is deemed approximately unbiased. Right: The same image, but only where the low resolution runs are approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

From the multiple low- and high-resolution spectra across all cosmologies, \cite{moran2023mira} obtained estimates for the precision of these spectra across the $n$ wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier $c$ for the increase in precision from the low- to high- resolution output. Given that the perturbation theory spectra is non-stochastic, we treat this as having a precision of $10^5$. We leverage these precision estimates within our model at each data type's unbiased wavenumber values to define $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, $\Lambda_{\ell} = \text{diag}\left(16(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, $\Lambda_h = \text{diag}\left(c(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$. Here, each precision matrix is diagonal and the diagonal elements consists of the precision values or zero, depending on whether the $k$ value is approximately unbiased for that data type. With this, we calculate a weighted average for each cosmology: $\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for the weighted average for a particular cosmology.

We will handle the data in two stages.  First, we will focus on 111 training cosmologies and obtain the best esimated spectrum for each.  Second, we will use these estimates to predict the spectra for six held-out cosmologies.

@@

From plots of the multiple low-resolution spectra, we find they vary smoothly about some mean $S$, suggesting that a dense covariance matrix $\Sigma_\varepsilon$ may outperform a diagonal covariance $\Lambda_\ell^{-1}$ for $\bar Y$. In previous work, we have modeled multiple candidates for $\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance $\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision values and subtracting a LOESS-smoothed average) performs the best. We specify this structure within $\Sigma_\varepsilon$ and refer readers to \cite{walsh2023bayesian} for a more detailed exposition. This yields $\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$, where $\Lambda_p, \Sigma_\ell^{-1}, \text{ and } \Lambda_h$ are all $n\times n$ and are 0 wherever the corresponding data type is biased.

We transform $X$ values to fall within $[0,1]$ and $Y$ so that the mean and variance are 0 and 1, respectively. This allows for a strightforward specification of priors, where the hyperparameters for each Mat\'ern covariance $\Sigma_S$ and $\Sigma_\ell$ each take a vague inverse-Gamma prior. Specifically:

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ following \cite{sauer2023active}:

\textcolor{cyan}{Annie, please include the most up-to-date prior specs here.}

@@

Upon integrating out $S$, we obtain $\bar Y \sim GP(\mu, \Sigma_S+\Sigma_\varepsilon)$. Additionally, this model yields the posterior distribution $S|\bar Y \sim N(m, C)$, where $C=\left(\Sigma_S^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1}$ and $m=C\left(\Sigma_\varepsilon^{-1}\bar Y+\Sigma_S^{-1}\mu\right)$. 
This distribution provides the posterior mean and corresponding uncertainty for $S$ within a given cosmology. Without this hierarchy, uncertainty would not be properly addressed; the weighted average would simply be nearly interpolated across the relatively dense set of $k$ values.

@@

Considering the Mira-Titan data, some nonstationarity is expected (e.g., due to baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$). To account for this, we incorporate a latent layer warping $W$ within the hierarchical model to incorporate nonstationary dynamics of the matter power spectra. When $W$ is a GP, this creates a deep Gaussian process (DGP) model \citep{damianou2013deep}. Additional latent layers could be considered, although one is sufficient for this dataset \citep{dunlop2018deep}.

@@

We have found that whether the prior mean of $W$ is set to 0 or $X$ \citep[which would indicate stationarity apriori, ][]{schmidt2003bayesian} does not have a substantial impact on the model fit. We construct this model in the compositional form, and also use MCMC to obtain full uncertainty for each of the model parameters \citep{sauer2023active}.

\textcolor{cyan}{Annie: want to add something here? We discuss the elliptical slice sampling of $W$, but otherwise it is plugged in to the methods of the previous section with very few changes.}

@@

We compare our DGP.HM model to competing models with a simulation study, in which case the true curve is known. Following \cite{moran2024dpc}, we consider two functions ($f_1$ and $f_2$) each with two variance specifications (settings ``4" and ``5", \textcolor{magenta}{we can consider additional homoskedastic cases in the appendix if necessary}).

@@

For the two variance settings, we consider matrices $\Sigma_4$ and $\Sigma_5$ to have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. For setting ``4", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``5", we we have $\tau^2=0.01$ and $\theta=0.05$, and additionally scale this covariance matrix by a vector of values $s$ so that $\Sigma_5 = s M s^T$, where $M$ is a Mat\'ern covariance matrix.

@@

The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates better performance. MSE and log score results are shown in Figures \ref{fig:sims_MSE} and \ref{fig:sims_logS} respectively; our DGP.HM model performs favorably against these current benchmark methods.

@@

In addition to simulated data, we investigate model fits the for different cosmological datasets (based on linear theory and/or runs from the Code for Anisotropies in the Microwave Background, CAMB). \textcolor{magenta}{We have looked at two sets of model runs, "LT" runs where there is an underlying ``true" value alongside 15 computer model runs, or the "CAMB" runs, where we only have a single output. The performance from the CAMB runs was better (with a 6-dimensional design space), and also did not require obtaining a posterior mean from multiple model runs. When looking at the CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error. Regarding these results, compared to the results for the 8-dimensional LT runs which had less impressive predictive performance, I'm not sure if we will include details on one or the other.}

@@

To being, we initialize our hyperparameter estimates across all cosmologies by modeling the first (M001) with 50,000 MCMC runs, and use these as initial values for all cosmologies. Specifically, we save the last MCMC iteration of $W$, $\theta_W$ and $\theta_S$ as $W_0$, $\theta_W0$ and $\theta_S0$ respectively and use each as starting values for all 111 training cosmology fits, where these are now fit with 10,000 MCMC runs (5,000 removed as burn-in and every 5th sample is kept, resulting in 1,000 posterior samples).

Here we can provide figures of the predicted mean and variances for a subset of the actual cosmologies. Figure \ref{fig:plot_fit} shows an example of the uncertainty for our posterior mean for the first (of 111) training cosmologies; the posterior mean is subtracted in order to visualize this uncertainty.

@@

This section details our second contribution - the use of the Bayesian DGP model of the previous section to predict the curve for held-out cosmologies.

@@

The different appendices from dissertation as necessary. Can also put details of dataset in supplementary material. \textcolor{magenta}{Fix references as necessary, e.g. Moran et al. 2024).}

>>>>

Leveraging these simulations allows for emulation and prediction of the matter power spectrum under varying specifications for eight different cosmological parameters (i.e., different cosmologies). Estimating and understanding the influence of these parameters on cosmic expansion are areas of active research. In addition to presenting the Mira-Titan simulation suite, \cite{moran2023mira} builds off of previous spectrum emulation \citep{lawrence2017mira} to provide the final emulator (Cosmic Emu) based on the full suite of 117 batches of simulations.

In this work, we expand on the previous emulation approaches and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical modeling and basis functions within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}. \textcolor{red}{Please note: this repository is currently private until we hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 117 different cosmologies. Each of these cosmologies is specified by 8 cosmological parameter values, which we will denote as $\theta \in \mathbb{R}^8$. Different simulation output exists for different values of a ninth value, the redshift; in this work, we fix this value at 0. For each cosmology, we have a batch of 18 simulated spectra: an inexpensive spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from a high resolution simulation ($y_h$).

@@

From the multiple low- and high-resolution spectra across all cosmologies, \cite{moran2023mira} obtained precision estimates of these spectra across the $n$ wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier $c$ for the increase in precision from the low- to high- resolution output. Given that the perturbation theory spectra is non-stochastic, we treat this as having a precision of $10^5$. We leverage these precision estimates within our model at each data type's unbiased wavenumber values to define $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, $\Lambda_{\ell} = \text{diag}\left(16(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, $\Lambda_h = \text{diag}\left(c(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$. Here, each precision matrix is diagonal and the diagonal elements consist of the precision values or zero, depending on whether the $k$ value is approximately unbiased for that data type. With this, we calculate a weighted average for each cosmology: $\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for the weighted average for a particular cosmology.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_data.png}
    \caption{Left: The perturbation theory, low resolution runs, and high resolution run for the first training cosmology. The weighted average of these is shown as a dashed line. Dotted lines at the bottom indicate where each data type is deemed approximately unbiased. Right: The same image, but restricted to $k$ values where the low resolution runs are approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

We will handle the data in two stages.  First, we will focus on 111 training cosmologies and obtain the best estimated spectrum for each.  Second, we will use these estimates to predict the spectra for six held-out cosmologies.

@@

From plots of the multiple low-resolution spectra (e.g., Figure \ref{fig:plot_data}), we find they vary smoothly about some mean $S$, suggesting that a dense covariance matrix $\Sigma_\varepsilon$ (accounting for spatial dependence) may outperform a diagonal covariance $\Lambda_\ell^{-1}$ for $\bar Y$. In previous work, we have modeled multiple candidates for $\Sigma_\varepsilon$ \citep{walsh2023bayesian} and found that a Mat\'ern covariance $\Sigma_\ell$ trained on the low-resolution runs (pre-scaled by the precision values and subtracting a LOESS-smoothed average) performs the best. We specify this structure within $\Sigma_\varepsilon$ and refer readers to \cite{walsh2023bayesian} for a more detailed exposition. This yields $\Sigma_\varepsilon=\left(\Lambda_p^{-1} + \Sigma_\ell + \Lambda_h^{-1}\right)^{-1}$, where $\Lambda_p, \Sigma_\ell^{-1}, \text{ and } \Lambda_h$ are all $n\times n$ and are 0 wherever the corresponding data type is biased.

We transform $X$ values to fall within $[0,1]$ and $Y$ so that the mean and variance are 0 and 1, respectively. This allows for a straightforward specification of priors, where the hyperparameters for the Mat\'ern covariance $\Sigma_S$ each take a vague prior. Specifically:

@@

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ following \cite{sauer2023active}: \textcolor{cyan}{Annie, please ensure the most up-to-date prior specs are shown above and here. Here is what I found from \texttt{fit\$settings} when using \texttt{deepgp v1.1.3}: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

Upon integrating out $S$, we obtain $\bar Y \sim \mathcal{N}_n(\mu, \Sigma_S+\Sigma_\varepsilon)$. Additionally, this model yields the posterior distribution $S|\bar Y \sim \mathcal{N}_n(m, C)$, where $C=\left(\Sigma_S^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1}$ and $m=C\left(\Sigma_\varepsilon^{-1}\bar Y+\Sigma_S^{-1}\mu\right)$. 
This distribution provides the posterior mean and corresponding uncertainty for $S$ within a given cosmology. 
% With this hierarchical model, we improve the efficiency of estimating $S|\bar{Y}$.

@@

Considering the Mira-Titan data, some nonstationarity is expected (e.g., due to baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$). To account for this, we incorporate a latent layer $W$ within the hierarchical model to warp locations of $X$ and incorporate nonstationary dynamics of the matter power spectra. When $W$ is a GP, this creates a deep Gaussian process (DGP) model \citep{damianou2013deep}. Additional latent layers could be considered, although one is sufficient for this dataset \citep{dunlop2018deep}.

@@

We have found that whether the prior mean of $W$ is set to 0 or $X$ \citep[which would indicate stationarity apriori,][]{schmidt2003bayesian} does not have a substantial impact on the model fit. We construct this model in the compositional form, and also use Markov Chain Monte Carlo (MCMC) to estimate and obtain full uncertainty for each of the model parameters \citep{sauer2023active}.

\textcolor{cyan}{Annie: want to add something here? For $\theta_w$, we also specify a Gamma prior of the form $\theta_w \sim G(\alpha=1.2, \beta=2)$. We discuss the elliptical slice sampling of $W$, but otherwise it is plugged in to the methods of the previous section with very few changes.}

@@

We compare DGP.HM to competing models with a simulation study where the underlying true curve is known. Following \cite{moran2024dpc}, we consider two functions ($f_1$ and $f_2$) each with two variance specifications (settings ``4" and ``5", \textcolor{magenta}{we can consider additional homoskedastic cases in the appendix if necessary}).

@@

For the two variance settings, we consider matrices $\Sigma_4$ and $\Sigma_5$ to have a Mat\'ern covariance function with a nugget of $10^{-8}$ and smoothness of 2.5. For setting ``4", we have $\tau^2=0.0225$ and $\theta=0.01$. For setting ``5", we we have $\tau^2=0.01$ and $\theta=0.05$, and additionally scale this covariance matrix by a vector of values $s$ so that $\Sigma_5 = s M s^T$ is nonstationary; here $M$ is a Mat\'ern covariance matrix with parameters $\tau^2$ and $\theta$.

@@

The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates better performance. Log score and MSE results are shown in Figures \ref{fig:sims_logS} and \ref{fig:sims_MSE} respectively; our DGP.HM model performs favorably against these current benchmark methods.

@@

In addition to simulated data, we investigate model fits the for different cosmological datasets (based on linear theory and/or runs from the Code for Anisotropies in the Microwave Background, CAMB). \textcolor{magenta}{We have looked at two sets of model runs, ``LT" runs where there is an underlying ``true" value alongside 15 computer model runs, or the ``CAMB" runs, where we only have a single output. The performance from the CAMB runs was better (with a 6-dimensional design space), and also did not require obtaining a posterior mean from multiple model runs. When looking at the CAMB runs on the $\Delta$ space, we see it generally is withing $\pm 5\%$ error. Regarding these results, compared to the results for the 8-dimensional LT runs which had less impressive predictive performance, I'm open to suggestions on whether to include details on one or the other.}

@@

To train our hierarchical model, we initialize our hyperparameter estimates across all cosmologies by modeling the first (M001) with 50,000 MCMC runs, and use these as initial values for all cosmologies. Specifically, we save the last MCMC iteration of $W$, $\theta_w$ and $\theta_S$ as $W_0$, $\theta_{w0}$ and $\theta_{S0}$ respectively and use each as starting values for all 111 training cosmology fits, where each cosmology is fit with 10,000 MCMC runs (5,000 removed as burn-in and every 5th sample is kept, resulting in 1,000 posterior samples).

From these samples, we can estimate and quantify the uncertainty within each cosmology's average and latent layer $W$. Figure \ref{fig:plot_fit} shows an example of the uncertainty for our posterior mean for the first (of 111) training cosmologies; the posterior mean is subtracted in order to visualize this uncertainty. Figure \ref{fig:plot_warp} provides elliptical slice samples of $W$ along with the average warping. Some departure from stationarity is estimated in the higher $X$ (i.e., scaled $k$) values.

@@

Upon obtaining the posterior means from each of the 111 training cosmologies, this section details our second contribution - the use of the Bayesian DGP.HM model to predict the curve for held-out cosmologies.

@@

The different appendices from dissertation as necessary. Can also put details of dataset in supplementary material. \textcolor{magenta}{Fix references as necessary, e.g. Moran et al. 2024.}
